### 2013-07-24 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding summary sections in cconclusion chapter

\chapter{Monte Carlo Synthetic Acceleration Methods}
\label{ch:stochastic_methods}
An alternative approach to approximate matrix inversion is to employ
Monte Carlo methods that sample a distribution with an expectation
value equivalent to that of the inverted operator. Such methods have
been in existence for decades with the earliest reference noted here a
manuscript published in 1950 by Forsythe and Leibler
\citep{forsythe_matrix_1950}. In their outline, Forsythe and Leibler
in fact credit the creation of this technique to J. Von Neumann and
S.M. Ulam some years earlier than its publication. In 1952 Wasow
provided a more formal explanation of Von Neumann and Ulam's method
\citep{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
\citep{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
\citep{spanier_monte_1969} present additional detail on this topic
using a collection of references from the 1950's and early
1960's.

Following this work, Halton presented a residual Monte Carlo scheme
that improved dramatically on the performance of these methods
\citep{halton_sequential_1962}. Recent work in this area has utilized
these residual Monte Carlo methods in particle transport
\citep{evans_residual_2003}, giving accelerated convergence over
traditional methods. As an expansion of Halton's work, Evans and
others have leveraged these methods in a synthetic acceleration scheme
that has been shown to be competitive with production Krylov methods
for radiation transport problems and improve performance significantly
over Halton's sequential method \citep{evans_monte_2012}.

@@

In a discussion of methods for solving linear systems, several
mathematical tools are useful in characterizing the qualities of the
linear system. Among the most useful are the \textit{eigenvalues} of
the matrix, $\sigma(\ve{A})$. We find these by solving the eigenvalue
problem:
\begin{equation}
  \ve{A} \ve{x} = \lambda \ve{x},\ \lambda \in \sigma(\ve{A})\:.
  \label{eq:eigenvalue_problem}
\end{equation}
By writing Eq~(\ref{eq:eigenvalue_problem}) in a different form,
\begin{equation}
  (\ve{A} - \lambda \ve{I})\ve{x} = 0 \:,
  \label{eq:eigenvalue_problem_2}
\end{equation}
and demanding that non-trivial solutions for $\ve{x}$ exist, it is
then required that $|\ve{A} - \lambda \ve{I}| = 0$. Expanding this
determinant yields a characteristic polynomial in terms of $\lambda$
with roots that form the set of eigenvalues, $\sigma(\ve{A})$. Each
component of $\sigma(\ve{A})$ can then be used to solve
Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
$\ve{x}$. The set of all permutations form the \textit{eigenvectors}
of $\ve{A}$. A quantity of particular interest that is computable from
the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral radius},
$\rho(\ve{A})$, defined by Saad \citep{saad_iterative_2003} as:
\begin{equation}
  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:.
  \label{eq:spectral_radius}
\end{equation}

%%---------------------------------------------------------------------------%%
\section{Neumann-Ulam Method}
\label{sec:mc_preliminaries}
We begin our discussion of Monte Carlo methods by seeking a solution
to Eq~(\ref{eq:linear_problem}). For a given linear operator $\ve{A}$,
we can use diagonal splitting in a similar manner as the stationary
method in Eq~(\ref{eq:linear_split_equation2}) to define the following
operator\footnote{It should be noted that non-diagonal splittings have
  been recently explored \citep{srinivasan_monte_2010} and have the
  potential to improve efficiency. However, it was observed in this
  work that this type of splitting did not improve performance in the
  asymptotic limit of $\rho(\ve{H}) \rightarrow 1$ for non-trivial
  problems.}:
\begin{equation}
  \ve{H} = \ve{I} - \ve{A}\:,
  \label{eq:linear_mc_iteration_matrix}
\end{equation}
such that we are solving the system:
\begin{equation}
  \ve{x} = \ve{H} \ve{x} + \ve{b}\:.
  \label{eq:richardson_split}
\end{equation}
We can then form an alternative representation for $\ve{A}^{-1}$ by
generating the \textit{Neumann series}:
\begin{equation}
  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
  \label{eq:neumann_series}
\end{equation}
which will converge if the spectral radius of $\ve{H}$ is less than
1. If we then apply this Neumann series to the right hand side of
Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
problem:
\begin{equation}
  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
  \label{eq:neumann_solution}
\end{equation}
An approximation of this summation by truncation will therefore lead
to an approximation of the solution. If we expand the summation with a
succession of matrix-vector multiply operations, we arrive at an
alternative perspective of this summation by considering the $i^{th}$
component of the solution vector:
\begin{equation}
  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
  \label{eq:expanded_neumann_solution}
\end{equation}
which can interpreted as a series of transitions between states,
\begin{equation}
 \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
 \rightarrow i_{k}\:,
  \label{eq:mc_walk_permutation}
\end{equation}
in $\ve{H}$ where $\nu$ is interpreted as a particular random walk
sequence permutation. We can generate these sequences of transitions
through Monte Carlo random walks by assigning them both a probability
and weight. As a reinterpretation of the iteration matrix, we then
form the \textit{Neumann-Ulam decomposition} of \ve{H}:
\begin{equation}
  \ve{H} = \ve{P} \circ \ve{W}\:,
  \label{eq:neumann_ulam_decomposition}
\end{equation}
where $\circ$ denotes the Hadamard product operation\footnote{The
  Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
  element-wise as $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the
transition probability matrix, and $\ve{W}$ denotes the transition
weight matrix. This decomposition, a generalization of Dimov's work
\citep{dimov_new_1998}, is an extension of the original Neumann-Ulam
scheme in that now a weight cutoff can be used to terminate a random
walk sequence and therefore truncate the Neumann series it is
approximating. The formulation of $\ve{P}$ and $\ve{W}$ will be
dependent on whether we choose a direct or adjoint Monte Carlo
sequence to estimate the state transitions in
Eq~(\ref{eq:expanded_neumann_solution}). In the direct method, we will
use the provided linear operator $\ve{A}$ to form the Neumann Ulam
decomposition while the adjoint method will use the adjoint linear
operator ($\ve{A}^T$ for real-valued systems) to form the
decomposition.

@@

%%---------------------------------------------------------------------------%%
\section{Adjoint Monte Carlo Method}
\label{sec:adjoint_mc}
Often a more useful form, an alternative to forward Monte Carlo matrix
inversion is the adjoint method. We begin by defining the linear
system adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
vectors respectively and $\ve{A}^T$ is the adjoint operator for
$\ve{A} \in \mathbb{R}^{N \times N}$. We can split this equation to
mirror Eq~(\ref{eq:richardson_split}):
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
  \label{eq:adjoint_split_system}
\end{equation}
As was required for convergence with the direct method using
Eq~(\ref{eq:richardson_split}), the spectral radius of $\ve{H}$ must
remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
therefore has the same spectral radius. By defining the following
inner product equivalence \citep{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
Using these definitions, we can derive an estimator from the adjoint
method that will also give the solution vector, $\ve{x}$. As with the
direct method, we can acquire the adjoint solution by forming the
Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
We expand this summation to again yield a series of transitions that
can be approximated by a Monte Carlo random walk sequence, this time
forming the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the solution to
Eq~(\ref{eq:linear_problem}). Here we have 2 unknowns, $\ve{y}$ and
$\ve{d}$, and therefore we require two constraints to close the
system. We use Eq~(\ref{eq:adjoint_vector_relation}) as the first
constraint and as a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_j\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
convenient outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_j \rangle = x_j \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source
and the adjoint solution using a delta function source, we recover one
component of the original solution.

@@

\subsection{Expected Value Estimator}
\label{subsec:expected_value_estimator}
In addition to the collision estimator, an additional estimator is
available due to the work of Okten \citep{okten_solving_2005} that
uses the method of expected values as a means to improve the Monte
Carlo estimate. As outlined by Spanier and Gelbard
\citep{spanier_monte_1969}, the method of expected values is a
deterministic averaging of events that may potentially occur in the
Monte Carlo random walk sequence. Okten applied this principle
directly to discrete Monte Carlo by forming the \textit{expected value
  estimator} for a random walk of $k$ events:
\begin{equation}
  X_{j}(\nu) = b_j + \sum_{m=0}^k W_m h_{j,i_m}\,
  \label{eq:expected_value_estimator}
\end{equation}
where now the contribution of the iteration matrix is
deterministically averaged at step $m$ over all potential states $j$
that may be reached from the current state $i_m$. Via Okten, the
estimator can be shown to be unbiased through a comparison to
the expected value estimator. We can first rewrite the summation in
Eq~(\ref{eq:expected_value_estimator}):
\begin{equation}
  X_{j}(\nu) = b_j + \sum_{m=0}^k \sum_{i=1}^N W_m
  \delta_{i_m,i} h_{ji}\,
  \label{eq:unbiased_eval_1}
\end{equation}
where $N$ is the number of states in the system. Immediately, we see
the collision estimator as defined by
Eq~(\ref{eq:adjoint_permutation_contribution}) and can therefore write
the expectation value as:
\begin{equation}
  E\{X_{j}\} = b_j + \sum_{i=1}^N E\{X_{i}\} h_{ji}\,
  \label{eq:unbiased_eval_2}
\end{equation}
which is equivalently is the $j^{th}$ component of
Eq~(\ref{eq:richardson_split}):
\begin{equation}
  E\{X_{j}\} = b_j + \sum_{i=1}^N x_{i} h_{ji}\,
  \label{eq:unbiased_eval_2}
\end{equation}
and is therefore and unbiased estimate. Compared to the collision
estimator, the expected value estimator provides additional
information at every step of the random walk, yielding potentially
better statistics with the same amount of transport
work. Conveniently, even if no Monte Carlo histories are computed, the
expected value estimator still deterministically computes the first
term of the Neumann Series, $\ve{H}^0\ve{b}$, whereas the collision
estimator will provide no information.

@@

%%---------------------------------------------------------------------------%%
\section{Sequential Monte Carlo}
\label{sec:sequential_mc}
The direct and adjoint Neumann-Ulam methods described are limited by a
convergence rate of $1/\sqrt{N}$ by the Central Limit Theorem where
$N$ is the number of random walk permutations. In 1962, Halton
presented a residual Monte Carlo method that moves towards exponential
convergence rates \citep{halton_sequential_1962} and further refined
his work some years later \citep{halton_sequential_1994}. Applications
of his work by the transport community have confirmed convergence
rates on the order of of $e^{-N}$ \citep{evans_residual_2003}. In much
the same way as the projection methods outlined in
Appendix~\ref{ch:linear_problem}, Halton's method, sequential Monte
Carlo, utilizes the adjoint Monte Carlo solver as a means of directly
reducing the elements residual vector. He proposed the following
iterative scheme as a solution to Eq~(\ref{eq:linear_problem})\:
\begin{subequations}
  \begin{gather}
    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
    \ve{A}\boldsymbol{\delta}^{k} = \ve{r}^{k}\:,\\
    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k}\:,
  \end{gather}
  \label{eq:sequential_monte_carlo}
\end{subequations}
where the correction $\boldsymbol{\delta}^k$ is computed by the
adjoint Monte Carlo method at each iteration. The merits of Halton's
approach are immediately visible in that we have now broken the
binding of the convergence rate to the Central Limit Theorem. Here,
the Monte Carlo solver is used to produce a correction from the
residual, analogous to using the residual to extract a correction from
the search subspace in a projection method. By doing this, the Monte
Carlo error is bound in the correction used to update the solution and
therefore does not explicitly manifest itself in the overall
convergence of the solution. The downside of such a method is that if
the solution guess is poor, then many iterations are required in order
to reach exponential converge as the Monte Carlo error (and therefore
the Central Limit Theorem) does dominate in this situation.

>>>>

\chapter{Monte Carlo Synthetic Acceleration Methods}
\label{ch:stochastic_methods}
An alternative approach to approximate matrix inversion is to employ
Monte Carlo methods that sample a distribution with an expectation
value equivalent to that of the inverted operator. Such methods have
been in existence for decades with the earliest reference noted here a
manuscript published in 1950 by Forsythe and Leibler
\cite{forsythe_matrix_1950}. In their outline, Forsythe and Leibler
in fact credit the creation of this technique to J. Von Neumann and
S.M. Ulam some years earlier than its publication. In 1952 Wasow
provided a more formal explanation of Von Neumann and Ulam's method
\cite{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
\cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
\cite{spanier_monte_1969} present additional detail on this topic
using a collection of references from the 1950's and early
1960's.

Following this work, Halton presented a residual Monte Carlo scheme
that improved dramatically on the performance of these methods
\cite{halton_sequential_1962}. Recent work in this area has utilized
these residual Monte Carlo methods in particle transport
\cite{evans_residual_2003}, giving accelerated convergence over
traditional methods. As an expansion of Halton's work, Evans and
others have leveraged these methods in a synthetic acceleration scheme
that has been shown to be competitive with production Krylov methods
for radiation transport problems and improve performance significantly
over Halton's sequential method \cite{evans_monte_2012}.

@@

In a discussion of methods for solving linear systems, several
mathematical tools are useful in characterizing the qualities of the
linear system. Among the most useful are the \textit{eigenvalues} of
the matrix, $\sigma(\ve{A})$. We find these by solving the eigenvalue
problem:
\begin{equation}
  \ve{A} \ve{x} = \lambda \ve{x},\ \lambda \in \sigma(\ve{A})\:.
  \label{eq:eigenvalue_problem}
\end{equation}
By writing Eq~(\ref{eq:eigenvalue_problem}) in a different form,
\begin{equation}
  (\ve{A} - \lambda \ve{I})\ve{x} = 0 \:,
  \label{eq:eigenvalue_problem_2}
\end{equation}
and demanding that non-trivial solutions for $\ve{x}$ exist, it is
then required that $|\ve{A} - \lambda \ve{I}| = 0$. Expanding this
determinant yields a characteristic polynomial in terms of $\lambda$
with roots that form the set of eigenvalues, $\sigma(\ve{A})$. Each
component of $\sigma(\ve{A})$ can then be used to solve
Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
$\ve{x}$. The set of all permutations form the \textit{eigenvectors}
of $\ve{A}$. A quantity of particular interest that is computable from
the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral radius},
$\rho(\ve{A})$, defined by Saad \cite{saad_iterative_2003} as:
\begin{equation}
  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:.
  \label{eq:spectral_radius}
\end{equation}

%%---------------------------------------------------------------------------%%
\section{Neumann-Ulam Method}
\label{sec:mc_preliminaries}
We begin our discussion of Monte Carlo methods by seeking a solution
to Eq~(\ref{eq:linear_problem}). For a given linear operator $\ve{A}$,
we can use diagonal splitting in a similar manner as the stationary
method in Eq~(\ref{eq:linear_split_equation2}) to define the following
operator\footnote{It should be noted that non-diagonal splittings have
  been recently explored \cite{srinivasan_monte_2010} and have the
  potential to improve efficiency. However, it was observed in this
  work that this type of splitting did not improve performance in the
  asymptotic limit of $\rho(\ve{H}) \rightarrow 1$ for non-trivial
  problems.}:
\begin{equation}
  \ve{H} = \ve{I} - \ve{A}\:,
  \label{eq:linear_mc_iteration_matrix}
\end{equation}
such that we are solving the system:
\begin{equation}
  \ve{x} = \ve{H} \ve{x} + \ve{b}\:.
  \label{eq:richardson_split}
\end{equation}
We can then form an alternative representation for $\ve{A}^{-1}$ by
generating the \textit{Neumann series}:
\begin{equation}
  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
  \label{eq:neumann_series}
\end{equation}
which will converge if the spectral radius of $\ve{H}$ is less than
1. If we then apply this Neumann series to the right hand side of
Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
problem:
\begin{equation}
  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
  \label{eq:neumann_solution}
\end{equation}
An approximation of this summation by truncation will therefore lead
to an approximation of the solution. If we expand the summation with a
succession of matrix-vector multiply operations, we arrive at an
alternative perspective of this summation by considering the $i^{th}$
component of the solution vector:
\begin{equation}
  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
  \label{eq:expanded_neumann_solution}
\end{equation}
which can interpreted as a series of transitions between states,
\begin{equation}
 \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
 \rightarrow i_{k}\:,
  \label{eq:mc_walk_permutation}
\end{equation}
in $\ve{H}$ where $\nu$ is interpreted as a particular random walk
sequence permutation. We can generate these sequences of transitions
through Monte Carlo random walks by assigning them both a probability
and weight. As a reinterpretation of the iteration matrix, we then
form the \textit{Neumann-Ulam decomposition} of \ve{H}:
\begin{equation}
  \ve{H} = \ve{P} \circ \ve{W}\:,
  \label{eq:neumann_ulam_decomposition}
\end{equation}
where $\circ$ denotes the Hadamard product operation\footnote{The
  Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
  element-wise as $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the
transition probability matrix, and $\ve{W}$ denotes the transition
weight matrix. This decomposition, a generalization of Dimov's work
\cite{dimov_new_1998}, is an extension of the original Neumann-Ulam
scheme in that now a weight cutoff can be used to terminate a random
walk sequence and therefore truncate the Neumann series it is
approximating. The formulation of $\ve{P}$ and $\ve{W}$ will be
dependent on whether we choose a direct or adjoint Monte Carlo
sequence to estimate the state transitions in
Eq~(\ref{eq:expanded_neumann_solution}). In the direct method, we will
use the provided linear operator $\ve{A}$ to form the Neumann Ulam
decomposition while the adjoint method will use the adjoint linear
operator ($\ve{A}^T$ for real-valued systems) to form the
decomposition.

@@

%%---------------------------------------------------------------------------%%
\section{Adjoint Monte Carlo Method}
\label{sec:adjoint_mc}
Often a more useful form, an alternative to forward Monte Carlo matrix
inversion is the adjoint method. We begin by defining the linear
system adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
vectors respectively and $\ve{A}^T$ is the adjoint operator for
$\ve{A} \in \mathbb{R}^{N \times N}$. We can split this equation to
mirror Eq~(\ref{eq:richardson_split}):
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
  \label{eq:adjoint_split_system}
\end{equation}
As was required for convergence with the direct method using
Eq~(\ref{eq:richardson_split}), the spectral radius of $\ve{H}$ must
remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
therefore has the same spectral radius. By defining the following
inner product equivalence \cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
Using these definitions, we can derive an estimator from the adjoint
method that will also give the solution vector, $\ve{x}$. As with the
direct method, we can acquire the adjoint solution by forming the
Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
We expand this summation to again yield a series of transitions that
can be approximated by a Monte Carlo random walk sequence, this time
forming the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the solution to
Eq~(\ref{eq:linear_problem}). Here we have 2 unknowns, $\ve{y}$ and
$\ve{d}$, and therefore we require two constraints to close the
system. We use Eq~(\ref{eq:adjoint_vector_relation}) as the first
constraint and as a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_j\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
convenient outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_j \rangle = x_j \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source
and the adjoint solution using a delta function source, we recover one
component of the original solution.

@@

\subsection{Expected Value Estimator}
\label{subsec:expected_value_estimator}
In addition to the collision estimator, an additional estimator is
available due to the work of Okten \cite{okten_solving_2005} that
uses the method of expected values as a means to improve the Monte
Carlo estimate. As outlined by Spanier and Gelbard
\cite{spanier_monte_1969}, the method of expected values is a
deterministic averaging of events that may potentially occur in the
Monte Carlo random walk sequence. Okten applied this principle
directly to discrete Monte Carlo by forming the \textit{expected value
  estimator} for a random walk of $k$ events:
\begin{equation}
  X_{j}(\nu) = b_j + \sum_{m=0}^k W_m h_{j,i_m}\,
  \label{eq:expected_value_estimator}
\end{equation}
where now the contribution of the iteration matrix is
deterministically averaged at step $m$ over all potential states $j$
that may be reached from the current state $i_m$. Via Okten, the
estimator can be shown to be unbiased through a comparison to
the expected value estimator. We can first rewrite the summation in
Eq~(\ref{eq:expected_value_estimator}):
\begin{equation}
  X_{j}(\nu) = b_j + \sum_{m=0}^k \sum_{i=1}^N W_m
  \delta_{i_m,i} h_{ji}\,
  \label{eq:unbiased_eval_1}
\end{equation}
where $N$ is the number of states in the system. Immediately, we see
the collision estimator as defined by
Eq~(\ref{eq:adjoint_permutation_contribution}) and can therefore write
the expectation value as:
\begin{equation}
  E\{X_{j}\} = b_j + \sum_{i=1}^N E\{X_{i}\} h_{ji}\,
  \label{eq:unbiased_eval_2}
\end{equation}
which is equivalently is the $j^{th}$ component of
Eq~(\ref{eq:richardson_split}):
\begin{equation}
  E\{X_{j}\} = b_j + \sum_{i=1}^N x_{i} h_{ji}\,
  \label{eq:unbiased_eval_2}
\end{equation}
and is therefore and unbiased estimate. Compared to the collision
estimator, the expected value estimator provides additional
information at every step of the random walk, yielding potentially
better statistics with the same amount of transport
work. Conveniently, even if no Monte Carlo histories are computed, the
expected value estimator still deterministically computes the first
term of the Neumann Series, $\ve{H}^0\ve{b}$, whereas the collision
estimator will provide no information.

@@

%%---------------------------------------------------------------------------%%
\section{Sequential Monte Carlo}
\label{sec:sequential_mc}
The direct and adjoint Neumann-Ulam methods described are limited by a
convergence rate of $1/\sqrt{N}$ by the Central Limit Theorem where
$N$ is the number of random walk permutations. In 1962, Halton
presented a residual Monte Carlo method that moves towards exponential
convergence rates \cite{halton_sequential_1962} and further refined
his work some years later \cite{halton_sequential_1994}. Applications
of his work by the transport community have confirmed convergence
rates on the order of of $e^{-N}$ \cite{evans_residual_2003}. In much
the same way as the projection methods outlined in
Appendix~\ref{ch:linear_problem}, Halton's method, sequential Monte
Carlo, utilizes the adjoint Monte Carlo solver as a means of directly
reducing the elements residual vector. He proposed the following
iterative scheme as a solution to Eq~(\ref{eq:linear_problem})\:
\begin{subequations}
  \begin{gather}
    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
    \ve{A}\boldsymbol{\delta}^{k} = \ve{r}^{k}\:,\\
    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k}\:,
  \end{gather}
  \label{eq:sequential_monte_carlo}
\end{subequations}
where the correction $\boldsymbol{\delta}^k$ is computed by the
adjoint Monte Carlo method at each iteration. The merits of Halton's
approach are immediately visible in that we have now broken the
binding of the convergence rate to the Central Limit Theorem. Here,
the Monte Carlo solver is used to produce a correction from the
residual, analogous to using the residual to extract a correction from
the search subspace in a projection method. By doing this, the Monte
Carlo error is bound in the correction used to update the solution and
therefore does not explicitly manifest itself in the overall
convergence of the solution. The downside of such a method is that if
the solution guess is poor, then many iterations are required in order
to reach exponential converge as the Monte Carlo error (and therefore
the Central Limit Theorem) does dominate in this situation.

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration}
\label{sec:mcsa}
Using the ideas of Halton, Evans and Mosher recently developed a Monte
Carlo solution method that was not prohibited severely by the quality
of the initial guess for the system \cite{evans_monte_2009} and later
applied it more rigorously as a solution mechanism for the radiation
diffusion equation \cite{evans_monte_2012}. With their new methods,
they achieved identical numerical results as conventional Krylov
solvers as well as comparable performance in both number of iterations
and CPU time. Their approach was instead to use residual Monte Carlo
as a synthetic acceleration for a stationary method. To derive this
method, we begin by splitting the operator in
Eq~(\ref{eq:linear_problem})
\begin{equation}
  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
  \label{eq:linear_split}
\end{equation}
With this we can then define the stationary method
\textit{Richardson's iteration} as:
\begin{equation}
  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
  \label{eq:richardsons_iteration}
\end{equation}
which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
solution error at the $k^{th}$ iterate relative to the true solution:
\begin{equation}
  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
  \label{eq:mcsa_error}
\end{equation}
Subtracting Eq~(\ref{eq:richardsons_iteration}) from
Eq~(\ref{eq:linear_split}) we get:
\begin{equation}
  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
  \label{eq:mcsa_setup_1}
\end{equation}
Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
\begin{equation}
  \begin{split}
    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
    \label{eq:mcsa_setup_2}
  \end{split}
\end{equation}
Using this, we define the following scheme that will converge in one
iteration if $\ve{A}$ is inverted exactly:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
  \end{gather}
  \label{eq:mcsa_setup_3}
\end{subequations}
However, $\ve{A}$ is only approximately inverted by our numerical
methods and therefore we instead pose an iterative scheme in which the
Monte Carlo solvers are used to invert the operator. The
\textit{Fixed-Point Monte Carlo Synthetic Acceleration} (MCSA) method
is defined as:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where a Neumann-Ulam Monte Carlo method is used to generate the
solution correction from the residual and Richardson's iteration in
the first step has been rewritten as a residual correction. Using
Monte Carlo in this way achieves the same effect as Halton's method,
decoupling its convergence rate from the overall convergence rate of
the method. Here, the approximate Monte Carlo solution is not driven
to a particular convergence as it merely supplies a correction for the
initial guess generated by Richardson's iteration. Rather, only a set
number of histories are required using the Neumann-Ulam method to
generate the correction. In addition, the fact that the scheme in
Eq~(\ref{eq:mcsa_setup_3}) will converge in one iteration if $\ve{A}$
is inverted exactly means that as more and more stochastic histories
are used to compute the correction and the error is reduced towards
zero, the number of iterations required for MCSA to converge should
decrease accordingly, thus accelerating the solution.
